Full-core analysis of thorium-fueled Molten Salt Breeder Reactor using the SERPENT 2 Monte Carlo code

### 2017-08-14 full-core_msbr_model.tex

The thermal spectrum \gls{MSBR} was designed to realize the promise of the 
thorium fuel cycle, which uses of natural thorium instead of enriched uranium.  
Thorium breeds fissile $^{233}$U and avoids uranium enrichment. The mixture of 
LiF-BeF$_2$-ThF$_4$-UF$_4$-PuF$_3$ has a melting point of $499^\circ$C, a low 
vapor pressure at operating temperatures, and good flow and heat transfer 
properties \cite{robertson_conceptual_1971}. The \gls{MSBR} complex geometry is 
challenging to describe in software input and usually require major geometric 
approximations\cite{park_whole_2015}.

@@

Fig.~\ref{fig:plan} shows the plan view of the whole-core configuration at the 
expected reactor operational level when both graphite control rods are fully 
inserted, and the safety rods are fully withdrawn. The safety rods only get 
inserted during an accident.  \begin{figure}[htbp!] % replace 't' with 'b' to 
        force it to be on the bottom
  \centering
  \vspace{-0.3em}
  \includegraphics[width=0.95\linewidth]{figure_2_1.png}
  \caption{Plan view of \gls{MSBR} core.}
  \vspace{-0.6em}
  \label{fig:plan}
\end{figure}
\FloatBarrier
Fig.~\ref{fig:elevation} shows the longitudinal section of the reactor. The 
violet color represents bare graphite, and the yellow represents fuel salt. The 
blue color shows Hastelloy-N, a material used for the plenum and vessel wall, 
and the black color is a void space. In this work, all figures of the core were 
generated using the built-in Serpent plotter.

\begin{figure}[htbp!] % replace 't' with 'b' to force it to be on the bottom
  \centering
  \includegraphics[width=\linewidth]{figure_2_2.}
  \caption{Elevation view of \gls{MSBR} core.}
  \label{fig:elevation}
\end{figure}

@@

The central portion, called Zone I, is made up of 1320 graphite elements, each 
$10.16$cm$\times10.16$cm$\times396.24$cm.
In Zone I, 13\% of the volume is fuel salt and 87\% is graphite. Zone I is 
composed of 1320 graphite cells and 4 channels for control rods : two for 
graphite rods which both regulate and shim during normal operation, and two
for backup safety rods to assure sufficient negative reactivity for emergency 
situations.

@@

\subsection{Core Zone II-B}

The undermoderated zone, Zone II, surrounds Zone I.
Combined with the bounding radial reflector, Zone II serves to diminish neutron 
leakage. This zone is formed of two kinds of elements: elements like those in 
Zone I with a larger channel diameter (Zone II-A), and radial
graphite slats (Zone II-B).

Zone II is Zone II 37\% salt by volume and each element has a fuel channel 
diameter of 6.604cm. It is divided into two different zones: Zone II-A and Zone 
II-B. The graphite elements for Zone II-A are prismatic. Zone II-B elements are 
rectangular slats spaced far enough apart to provide the 0.37 fuel salt volume 
fraction. Fig.~\ref{fig:zone2B} additionally shows the 5.08cm-wide annular 
space between the core graphite and the radial reflector graphite. The annulus 
contains 100\% fuel salt and serves to reduce the damage flux at the internal 
surface of the graphite reflector blocks. The reactor Zone II-B graphite 
5.08cm-thick slats vary in the radial dimension (average width is 26.67cm) but 
are reconstructed without any approximation. From the ORNL report 
\cite{robertson_conceptual_1971}, the suggested design of Zone II-B has 8 
irregularly-shaped graphite elements every 45$^\circ$ as well as salt channels. 
These graphite elements were simplified into right-circular cylindrical shapes 
with central channels. This is the only simplification made to the \gls{MSBR} 
conceptual geometry in this work.
\FloatBarrier
\begin{figure}[htp!] % replace 't' with 'b' to force it to be on the bottom
  \centering
  \includegraphics[width=\linewidth]{figure_2_4.png}
  \caption{Zone I (left) and Zone II-A (right) elements.}
  \label{fig:zone12A}
\end{figure}
\subsection{Material composition}
The fuel salt, the reactor graphite, and the modified HastelloyN are materials 
unique of the \gls{MSBR} and were created at \gls{ORNL}. The initial fuel salt 
loading composition is LiF-BeF$_2$-ThF$_4$-$^{233}$UF$_4$ (71.8-16-12-0.2 mole 
\%). The lithium in the molten salt fuel is a fully enriched $^{7}$Li because 
$^{6}$Li is a very strong neutron poison and becomes tritium upon neutron 
capture. For cross section generation, ENDF/B-VII was employed 
\cite{chadwick_endf/b-vii.0:_2006}. The specific temperature was fixed for each 
material to correctly model the Doppler-broadening of resonance peaks when 
Serpent generate problem-oriented nuclear data library.
%The isotope composition of each material at the initial state was described in 
%detail in the MSBR conceptual design study \cite{robertson_conceptual_1971} 
%and has been applied to Serpent model without any modification. %
\begin{figure}[htbp!] % replace 't' with 'b' to force it to be on the bottom
  \centering
  \includegraphics[width=0.96\linewidth]{figure_2_5.png}
  \caption{Plan view that includes Zone I, II-A, and II-B elements.}
  \label{fig:zone2B}
\end{figure}
\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{R}
This section presents calculation results, such as the effective multiplication 
factor for whole core, neutron flux spectrum, and temperature reactivity 
coefficients. The normalized neutron flux distribution is calculated for the 
whole core using continuous-energy nuclear data.
The temperature coefficients for both fuel salt solution and reactor graphite 
are computed by comparing effective multiplication factors for two temperatures 
in the working range.

\subsection{Neutron spectrum}
Fig.~\ref{fig:spectrum} demonstrates the normalized neutron flux spectrum for 
the whole core in the energy range from $10^{-9}$ to $10$ MeV. The results show 
close fit with the MCNP simulation \cite{park_whole_2015}, especially in 
thermal energy range.  \begin{figure}[h!] % replace 't' with 'b' to force it to 
        be on the bottom
  \centering
  \includegraphics[width=1.05\linewidth]{figure_3_1.png} \caption{Neutron flux 
  spectrum of \gls{MSBR} for MCNP6 and Serpent 2 model.}
  \label{fig:spectrum}
\end{figure}
It is important to obtain the epithermal and thermal spectrum to produce 
$^{233}$U from $^{232}$Th because radiative capture cross section of thorium 
monotonically decreases from $10^{-10}$ MeV to $10^{-5}$ MeV. Hardening the 
spectrum tends to significantly increase resonance absorption in thorium and 
decrease the absorptions in fissile and construction materials. Thus, a 
signficant amount fissile material will be needed to make the reactor critical.

\subsection{Effective multiplication factor}
Table~\ref{tab:keff} shows the effective multiplication factor for both MCNP6 
and Serpent 2 whole core models. The factor obtained using Serpent 2 is 300 pcm 
lower than that obtained by Park \emph{et al.} using MCNP6 
\cite{park_whole_2015}. Standard deviations are 5 and 9 pcm, respectively. The 
discrepancy is likely due to simplificiations to the Zone II geometry model 
used in Park \emph{et al.}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\captionsetup[table]{
  labelsep = newline,
  name = TABLE, justification=justified,
  singlelinecheck=false,%%%%%%% a single line is centered by default
  labelsep=colon,%%%%%%
  skip = \medskipamount}
\begin{table}[h!]
%\centering
\caption{Effective multiplication factor of whole core model.}
\begin{tabular}{p{0.15\linewidth} p{0.3\linewidth} p{0.3\linewidth}} \toprule
      & Serpent2      & MCNP6 \cite{park_whole_2015}          \\ \midrule
K$_{eff}$  & 1.00389$\pm$0.00005 & 1.00736$\pm$0.00009
\\
\bottomrule
\end{tabular}
  \label{tab:keff}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Temperature effect of reactivity}
Table~\ref{tab:tcoef} shows temperature effects on reactivity calculated in 
this work as compared to both \cite{park_whole_2015} and 
\cite{robertson_conceptual_1971}. Uncertainty for each temperature coefficient 
also appears in Table~\ref{tab:tcoef}. The main physical principle underlying 
the reactor temperature feedback is an expansion of matter when it is heated.  
When the fuel salt temperature increases, the density of the salt decreases, 
but at the same time, the total volume of fuel salt in the core remains 
constant because it is bounded by the graphite. When the reactor graphite 
temperature grows, the density of graphite declines creating additional space 
for fuel salt. To determine temperature coefficients, the cross-section 
temperatures for fuel and moderator were changed from 900K to 1200K. Three 
different cases were considered:
\begin{enumerate}  \item Temperature of fuel salt rising from 900K to 1200K.
\item Temperature of graphite rising from 900K to 1200K.  \item Whole reactor 
        temperature rising from 900K to 1200K.
\end{enumerate}

>>>>

The thermal spectrum \gls{MSBR} was designed to realize the promise of the 
thorium fuel cycle, which uses of natural thorium instead of enriched uranium.  
Thorium breeds fissile $^{233}$U and avoids uranium enrichment. The mixture of 
LiF-BeF$_2$-ThF$_4$-UF$_4$-PuF$_3$ has a melting point of $499^\circ$C, a low 
vapor pressure at operating temperatures, and good flow and heat transfer 
properties \cite{robertson_conceptual_1971}. The \gls{MSBR} complex geometry is 
challenging to describe in software input and usually require major geometric 
approximations \cite{park_whole_2015}.

@@

Fig.~\ref{fig:plan} shows the plan view of the whole-core configuration at the 
expected reactor operational level when both graphite control rods are fully 
inserted, and the safety rods are fully withdrawn. The safety rods only get 
inserted during an accident.

Fig.~\ref{fig:elevation} shows the longitudinal section of the reactor. The 
violet color represents bare graphite, and the yellow represents fuel salt. The 
blue color shows Hastelloy-N, a material used for the plenum and vessel wall, 
and the black color is a void space. In this work, all figures of the core were 
generated using the built-in Serpent plotter.

\begin{figure}[hbp!] % replace 't' with 'b' to 
  \centering
  \vspace{-0.3em}
  \includegraphics[width=0.95\linewidth]{figure_2_1.png}
  \caption{Plan view of \gls{MSBR} core.}
  \vspace{-0.6em}
  \label{fig:plan}
\end{figure}
\FloatBarrier

\begin{figure}[htbp!] % replace 't' with 'b' to force it to be on the bottom
  \centering
  \includegraphics[width=\linewidth]{figure_2_2.png}
  \caption{Elevation view of \gls{MSBR} core.}
  \label{fig:elevation}
\end{figure}

@@

The central portion, called Zone I, is made up of 1320 graphite elements, each 
$10.16$cm$\times$10.16cm$\times$396.24cm.
In Zone I, 13\% of the volume is fuel salt and 87\% is graphite. Zone I is 
composed of 1320 graphite cells and 4 channels for control rods: two for 
graphite rods which both regulate and shim during normal operation, and two
for backup safety rods to assure sufficient negative reactivity for emergency 
situations.

@@

\subsection{Core Zone II}
The undermoderated zone, Zone II, surrounds Zone I.
Combined with the bounding radial reflector, Zone II serves to diminish neutron 
leakage. This zone is formed of two kinds of elements: elements like those in 
Zone I with a larger channel diameter (Zone II-A), and radial
graphite slats (Zone II-B).

Zone II is Zone II 37\% salt by volume and each element has a fuel channel 
diameter of 6.604cm. It is divided into two different zones: Zone II-A and Zone 
II-B. The graphite elements for Zone II-A are prismatic. Zone II-B elements are 
rectangular slats spaced far enough apart to provide the 0.37 fuel salt volume 
fraction. Fig.~\ref{fig:zone2B} additionally shows the 5.08cm-wide annular 
space between the core graphite and the radial reflector graphite. The annulus 
contains 100\% fuel salt and serves to reduce the damage flux at the internal 
surface of the graphite reflector blocks. The reactor Zone II-B graphite 
5.08cm-thick slats vary in the radial dimension (average width is 26.67cm) but 
are reconstructed without any approximation. From the ORNL report 
\cite{robertson_conceptual_1971}, the suggested design of Zone II-B has 8 
irregularly-shaped graphite elements every 45$^\circ$ as well as salt channels. 
These graphite elements were simplified into right-circular cylindrical shapes 
with central channels. This is the only simplification made to the \gls{MSBR} 
conceptual geometry in this work.
\FloatBarrier
\begin{figure}[t] % replace 't' with 'b' to force it to be on the bottom
  \centering
  \includegraphics[width=\linewidth]{figure_2_4.png}
  \caption{Zone I (left) and Zone II-A (right) elements.}
  \label{fig:zone12A}
\end{figure}
\FloatBarrier
\subsection{Material composition}
The fuel salt, the reactor graphite, and the modified HastelloyN are materials 
unique of the \gls{MSBR} and were created at \gls{ORNL}. The initial fuel salt 
loading composition is LiF-BeF$_2$-ThF$_4$-$^{233}$UF$_4$ (71.8-16-12-0.2 mole 
\%). The lithium in the molten salt fuel is a fully enriched $^{7}$Li because 
$^{6}$Li is a very strong neutron poison and becomes tritium upon neutron 
capture. For cross section generation, ENDF/B-VII was employed 
\cite{chadwick_endf/b-vii.0:_2006}. The specific temperature was fixed for each 
material to correctly model the Doppler-broadening of resonance peaks when 
Serpent generate problem-oriented nuclear data library.
%The isotope composition of each material at the initial state was described in 
%detail in the MSBR conceptual design study \cite{robertson_conceptual_1971} 
%and has been applied to Serpent model without any modification. %
\begin{figure}[htbp!] % replace 't' with 'b' to force it to be on the bottom
  \centering
  \includegraphics[width=0.89\linewidth]{figure_2_5.png}
  \caption{Plan view that includes Zone I, II-A, and II-B elements.}
    \vspace{-0.6em}
  \label{fig:zone2B}
\end{figure}
\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
This section presents calculation results, such as the effective multiplication 
factor for whole core, neutron flux spectrum, and temperature reactivity 
coefficients. The normalized neutron flux distribution is calculated for the 
whole core using continuous-energy nuclear data.
The temperature coefficients for both fuel salt solution and reactor graphite 
are computed by comparing effective multiplication factors for two temperatures 
in the working range.

\subsection{Neutron spectrum}
Fig.~\ref{fig:spectrum} demonstrates the normalized neutron flux spectrum for 
the whole core in the energy range from $10^{-9}$ to $10$ MeV. The results show 
close fit with the MCNP simulation \cite{park_whole_2015}, especially in 
thermal energy range.  
\begin{figure}[b!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=1.05\linewidth]{figure_3_1.png} \caption{Neutron flux 
  spectrum of \gls{MSBR} for MCNP6 and Serpent 2 model.}
  \label{fig:spectrum}
\end{figure}
It is important to obtain the epithermal and thermal spectrum to produce 
$^{233}$U from $^{232}$Th because radiative capture cross section of thorium 
monotonically decreases from $10^{-10}$ MeV to $10^{-5}$ MeV. Hardening the 
spectrum tends to significantly increase resonance absorption in thorium and 
decrease the absorptions in fissile and construction materials. Thus, a 
signficant amount fissile material will be needed to make the reactor critical.

\subsection{Effective multiplication factor}
Table~\ref{tab:keff} shows the effective multiplication factor for both MCNP6 
and Serpent 2 whole core models. The factor obtained using Serpent 2 is 300 pcm 
lower than that obtained by Park \emph{et al.} using MCNP6 
\cite{park_whole_2015}. Standard deviations are 5 and 9 pcm, respectively. The 
discrepancy is likely due to simplificiations to the Zone II geometry model 
used in Park \emph{et al.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\captionsetup[table]{
  labelsep = newline,
  name = TABLE, justification=justified,
  singlelinecheck=false,%%%%%%% a single line is centered by default
  labelsep=colon,%%%%%%
  skip = \medskipamount}
\begin{table}[h!]
%\centering
\caption{Effective multiplication factor of whole core model.}
\begin{tabular}{p{0.15\linewidth} p{0.3\linewidth} p{0.3\linewidth}} \toprule
      & Serpent2      & MCNP6 \cite{park_whole_2015}          \\ \midrule
K$_{eff}$  & 1.00389$\pm$0.00005 & 1.00736$\pm$0.00009
\\
\bottomrule
\end{tabular}
  \label{tab:keff}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Temperature effect of reactivity}
Table~\ref{tab:tcoef} shows temperature effects on reactivity calculated in 
this work as compared to both \cite{park_whole_2015} and 
\cite{robertson_conceptual_1971}. Uncertainty for each temperature coefficient 
also appears in Table~\ref{tab:tcoef}. The main physical principle underlying 
the reactor temperature feedback is an expansion of matter when it is heated.  
When the fuel salt temperature increases, the density of the salt decreases, 
but at the same time, the total volume of fuel salt in the core remains 
constant because it is bounded by the graphite. When the reactor graphite 
temperature grows, the density of graphite declines creating additional space 
for fuel salt. To determine temperature coefficients, the cross-section 
temperatures for fuel and moderator were changed from 900K to 1200K. Three 
different cases were considered:
\begin{enumerate}  \item Temperature of fuel salt rising from 900K to 1200K.
\item Temperature of graphite rising from 900K to 1200K.  \item Whole reactor 
        temperature rising from 900K to 1200K.
\end{enumerate}
